### 2019-12-23 paper/educ_work_over_time.tex
Expand discussion of female labor supply theory and findings

Even as the level of female education has increased the female labor force participation 
has decreased over time [should this be by level of education?].

[Discuss recent papers on labor supply in India]

To place my results in context, Figure XX shows the percent of married women who are 
working at the time of the survey by age group.%
\footnote{
This question is the only question on labor force participation consistently available 
across all four surveys.
Since the question is asked only of married women it is possible that the overall
female labor force participation might have developed differently, especially since
many young, unmarried women have begun working in, for example, the business process 
outsourcing industry [TK cite Jensen 2012 QJE].
}

The declines over the the four rounds of the NFHS are in line with what previous
research has found using other data sets [TK cite Bhargava 2018 and 2019, Chatterjee 2018].
Finding a decline in labor force participation in the beginning of development is
in line with the hypothesis of a U-shaped labor force participation as a country
develop [TK cite Goldin 1994, Chatterjee 2018].

The theories on the relationship between education and birth spacing often works 
through the changes in the opportunity cost of time. 
If women are unlikely to work, this makes the predictions that are based on
opportunity cost less relevant. 
instead, theories that emphasize income effects or effects of spacing become
more relevant.

>>>>

Even as the level of female education has increased, the female labor force participation 
in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Afridi2018,Bhargava2018, Chatterjee2018, Bhargava2019}.
A decline in female labor force participation in the beginning of development is
consistent with the hypothesis of a U-shaped labor force participation as a country
develop \citep{Goldin1994}.
India's female labor force participation is, however, lower than most other
countries and more in line with countries in the Middle East and North Africa and
does not yet show any signs of increasing.
\citep{Klasen2015,Chatterjee2018}.

Theory predicts that increasing women's wage leads to higher labor force participation, 
while both a higher male income and social restrictions on working, possibly in 
specific industries or jobs, reduce female labor supply \citep{Goldin1994}.
India's economy has grown substantially since the mid-seventies with
an average annual growth rate of about 5.5 percent from 1978 to 2004 \citet{Bosworth2008}.
As a result, real wages for both men and women have almost doubled between 1987 and 2011 
\citep{Klasen2015}.
Despite the increases in wages and the higher female education, the mean male wage is
still close to 70 percent higher than the mean female wage \citep{Bhargava2018}.
Furthermore, the development in women's wage is not uniform across education levels.
The real wages for women with middle school education or below has converged, 
for women with secondary education, the change over time depends on the sample used
with some showing a relative strong decline and others a modest increase,
and women with more than secondary education experienced an increased in their real 
wage, especially in cities \citep{Klasen2015,Bhargava2018}.

After controlling for household characteristics and husband income, there is
still a U-shaped relationship between female education and labor supply with the
lowest labor force participation for women with secondary education \citep{Chatterjee2018}.
There are two possible explanations for this effect, both directly related to the strong
pro-male bias in India.
First, it becomes less socially acceptable for women to work in manual labor as their
education increases, but women are still mostly excluded from white-collar
employment despite the strong economic growth \citep{Klasen2015,Chatterjee2018}.
Second, with increasing education, women's productivity at home increases, especially
in the production of offspring human capital, and with access to sex selective abortions
leading to relatively more boys being born and an increasing return to male education 
there can be an increased demand for women with more education, even though they do not 
participate in the labor market \citep{Behrman1999}.

To place my results in context, Figure XX shows the percent of married women who are 
working at the time of the survey by age group.
This question is the only question on labor force participation consistently available 
across all four surveys, and because the question refers to currently working the 
percentages will be lower than what other studies have found.
Furthermore, since the question is asked only of married women, the overall female labor 
force participation might have developed differently, especially since many young, 
unmarried women have begun working in, for example, the business process 
outsourcing industry, which, in turn, has increased girls' schooling investments
\citep{Jensen2012}.

Women in their twenties have the lowest labor force participation, consistent with
that age being the prime age bearing period.
As a result, there is little change in the percent women who currently work, except
that rural women with no education has seen a 15 percentage points decline.

The theories on the relationship between education and birth spacing often works 
through the changes in the opportunity cost of time. 
If women are unlikely to work, this makes the predictions that are based on
opportunity cost less relevant. 
Instead, theories that emphasize income effects or effects of spacing on child
outcomes become more relevant.
